Approximating Likelihood Ratios with Calibrated Discriminative Classifiers

### 2015-06-03 tex/parametrized-learning-aoas.tex
updating related work for Neal

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}

@@

\begin{frontmatter}
%\title{Approximating Generalized Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
\title{Approximating  Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
%\title{Approximating Likelihood Ratio Tests with Calibrated Classifiers}
%\title{A Sample Document\thanksref{T1}}
\runtitle{Approximating the Likelihood Ratio with Classifiers}
%\thankstext{T1}{Footnote to the title with the ``thankstext'' command.}

@@

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
In many areas of science, generalized likelihood ratio tests  are established tools for statistical inference. 
It is increasingly common that a generative model is used to describe complex processes that tie 
parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical. Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for training and calibration.

%We demonstrate how discriminative classifiers can be used to approximate 
%generalized likelihood ratio tests over high-dimensional data when a generative model 
%for the data is available for training and calibration.  
\end{abstract}

@@

In many areas of science, (generalized) likelihood ratio tests  are established tools for statistical inference~\cite{Wilks1938}. 
It is increasingly common that a generative model is used to describe complex processes that tie 
parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical. Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for training and calibration.

As a concrete example, consider searches for new particles at the Large Hadron Collider. 
The ATLAS and CMS experiments have published several hundred papers where the 
final statistical result was formulated as a hypothesis test or confidence interval 
using a generalized likelihood ratio test~\cite{Cowan:2010js}. This includes
the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa} and 
subsequent measurement of its properties.

The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
that discriminates between a hypothesized process of interest (labeled \textit{signal}) and various other processes 
(labeled \textit{background}). To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers. Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\cite{Hocker:2007ht} are commonly used. Recently, there has been progress in using deep networks~\cite{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\cite{HepML}, the largest Kaggle challenge in history. 
%The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.

%Collisions are usually selected with supervised learning methods 
While classification accuracy can lead to optimal approaches for simple hypothesis tests~\cite{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters. As noted in ~\cite{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
A fundamental distinction is that evaluating the loss for classification is a per-event operation, while evaluating the loss for a mass measurement (and other estimation problems) is a per-experiment operation involving a dataset with many events. They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process. 
%This paper shows that (to the extent a generalized likelihood ratio test in the full feature space is close to  optimal) one can achieve optimal results with a  specific per-event procedure by reformulating the target likelihood ratio in terms of a per-event classification followed by an specific calibration process.

Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit. These asymptotic properties, the difficulty in specifying a prior over the parameters needed to evaluate a Bayes risk, and the vast literature associated to likelihood ratio tests has led many users of such tests to not be explicit about the loss function they wish to minimize for their experiment. Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself. The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss. This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process. Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.

@@

We use the following notation:
\begin{itemize}
 \item $x$: a vector of features for an event
 \item $D$: a dataset of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $p(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
\item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ \end{itemize}
We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.

@@

In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio 
\begin{equation}
T(D) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
\end{equation}
is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density 
$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that 
can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
high energy physics where the simulation of particle detectors can only be done in the `forward mode'. This same setting has been considered by \cite{ClaytonScott} and \cite{JMLR:v14:tong13a}.

@@

While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire dataset $D$, we are able to reformulate it such that the machine learning problem is a per-event classification problem. This follows from the fact that we assume the $x_e$ to be i.i.d.

@@

Vast literature exists around generative and discriminative classifiers~\cite{AndrewY.Ng}. Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule. In contrast, discriminative classifiers model the posterior $p(y|x)$ directly. For classification tasks, one then thresholds on $p(y|x)$. In both cases this description in terms of a posterior requires a prior distribution for $p(y)$, which is either modeled explicitly or learned from the training data. 
This familiar formulation of classification may lead to some confusion in the setting of the current work.

The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory, able to make predictions about $x$ and being motivated by domain-specific considerations. For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$.  
Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
%The point here is to make the distinction of $p(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.

The second possible source of confusion is that 
we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$. 
Instead, we are interested in the approximation of the per-experiment likelihood ratio, which can be used to calculate p-values defined by $P(T(D) > k |\theta)$.
%
%%the likelihood ratio $T(D)$ is aimed at tests based on the entire dataset; 
%we are not interested in thresholded classification on individual events $x_e$.  
%Knowing that classifier scores are often poorly calibrated is an immediate concern.

@@

Lastly, in the setting of frequentist hypothesis tests, we do not have a prior $\pi(\theta)$. 
While we can use the generative models to produce training data $(x,y=0)$ generated 
from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$, the relative mix $p(y)$ 
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
\end{equation}
which is monotonic with the desired likelihood ratio $p(x|\theta_1)/p(x|\theta_0)$.
Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.  Similar points have been made by~\cite{ClaytonScott}.

@@

It is not obvious that if we are working with the distributions $p(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators. 
Fortunately, there is a construction based on $p(s|\theta)$ that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use Theorem~1 to reformulate the maximum likelihood estimate
\begin{equation}
\hat{\theta} = \argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} = \argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
\end{equation}
It is important that we include the denominator $p(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

@@

We are left with the practical question of how to train a family of discriminative classifiers parametrized by $\theta_0$ and $\theta_1$, the 
parameters associated to the null and alternate hypotheses, respectively. While this could be done independently
for all $\theta_0$ and $\theta_1$, it is desirable and convenient to have a smooth evolution of the classification score as a function of the parameters. Thus, we anticipate a single learning stage based on training data with input $(x, \theta_0, \theta_1)_i$ and target $y_i$. Somewhat unusually, the unknown values of the parameters are taken
as input to the classifier, and are latent variables whose values will be specified via the enveloping (generalized) likelihood ratio test. We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
\begin{algorithm}[ht]
\caption{Training of the parametrized classifier.}\label{alg:training}
\begin{algorithmic}
\STATE initialize trainingData = \{\}
\FOR{ $\theta_0$ in $\Theta$ }
	\FOR{ $\theta_1$ in $\Theta$ }
		\STATE generate $x_i^0 \sim p(x|\theta_0)$
		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
		\STATE generate $x_i^1 \sim p(x|\theta_1)$
		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
	\ENDFOR
\ENDFOR
\STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
\end{algorithmic}
\end{algorithm}%\vspace{-1.5em}

@@

In most settings that make use of likelihood ratio tests, the likelihood is based directly on some approximation of density for the observed data via $\hat{f}(x|\theta)$.  In general, approximating the density $\hat{f}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{g}(\hat{s}|\theta)$.  In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$. In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$. A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\cite{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

@@

In HEP we are often searching for some 
class of events, generically referred to as \textit{signal}, in the presence of a separate class 
of \textit{background} events.  For each event we measure some quantities $x$ that have corresponding distributions 
$p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing 
uncertainties in the underlying physics prediction or response of the measurement device. The 
total model is a mixture of the signal and background, and $\mu$ is the mixture coefficient associate 
to the signal component. 
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\cite{Cowan:2010js}. This approach was used for the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.

%Generalized likelihood ratio tests are used widely in HEP ~\cite{Cowan:2010js}, most notably in the discovery of the Higgs boson~\cite{Aad:2012tfa,Chatrchyan:2012ufa}.

@@

This formalism represents is a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure. The work of  ~\cite{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed. This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.

Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\cite{Volobouev:2011vb}. In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method. Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time per event to calculate the likelihood ratio $q(x)$. The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation. It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage. In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\cite{Aaltonen:2010yz}).

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\cite{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\cite{Allanach:2000kt}.  Both of these examples involve datasets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters. 
%The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
%Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\cite{Bolognesi:2012mm}.

\subsection{Related work}

%Refs to include:
%\cite{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,}

>>>>

\DeclareMathOperator*{\argmin}{arg\,min}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\citek}[1]{[\cite{#1}]}

@@

\begin{frontmatter}
%\title{Approximating Generalized Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
%\title{Approximating  Likelihood Ratio Tests with Calibrated Discriminative Classifiers}
\title{Approximating  Likelihood Ratios with Calibrated Discriminative Classifiers}
%\title{Approximating Likelihood Ratio Tests with Calibrated Classifiers}
%\title{A Sample Document\thanksref{T1}}
\runtitle{Approximating Likelihood Ratios with Classifiers}
%\thankstext{T1}{Footnote to the title with the ``thankstext'' command.}

@@

\begin{abstract}%   <- trailing '%' for backward compatibility of .sty file
In particle physics likelihood ratio tests are established tools for statistical inference. 
These tests are complicated by the fact that computer simulators are used as a generative model for 
the data, but they do not provide a way to evaluate the likelihood function for a given
observation. 
We demonstrate how discriminative classifiers can be used to approximate the likelihood function when 
a generative model for the data is available for training and calibration.  
\end{abstract}

@@

The likelihood function is the central object that summarizes the information from an experiment needed for inference of model parameters. The likelihood function is key to Bayesian inference, and many areas of science that report the results of classical hypothesis tests or confidence intervals use the (generalized or profile) likelihood ratio as a test statistic.
%  are established tools for statistical inference~\citek{Wilks1938}. 
It is increasingly common that a simulator (or generative model) is used to describe complex processes that tie 
parameters $\theta$ of an underlying theory and measurement apparatus to high-dimensional observations $x$. 
Directly evaluating the likelihood ratio in these cases is often not possible or is computationally impractical. Here we demonstrate how discriminative classifiers can be used to construct equivalent likelihood ratio tests when 
a generative model for the data is available for training and calibration.

As a concrete example, consider searches for new particles at the Large Hadron Collider. 
The ATLAS and CMS experiments have published several hundred papers where the 
final statistical result was formulated as a hypothesis test or confidence interval 
using a generalized  likelihood ratio test~\citek{Cowan:2010js}. This includes
the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa} and 
subsequent measurement of its properties.

The bulk of the likelihood ratio tests at the LHC are based on the distribution of a single event-level feature
that discriminates between a hypothesized process of interest (labeled \textit{signal}) and various other processes 
(labeled \textit{background}). To improve the statistical power of these tests, hundreds of these searches have utilized supervised learning to train discriminative classifiers. Within High Energy Physics (HEP) libraries such as TMVA that implement conventional techniques like the multi-layer perceptron and boosted decision trees~\citek{Hocker:2007ht} are commonly used. Recently, there has been progress in using deep networks~\citek{Baldi:2014kfa} and a NIPS workshop synthesizing the lessons learned during HiggsML~\citek{HepML}, the largest Kaggle challenge in history. 
%The use of classifiers is effective, but an intermediate step not obviously connected to the target likelihood ratio test.

%Collisions are usually selected with supervised learning methods 
While classification accuracy can lead to optimal approaches for simple hypothesis tests~\citek{Dempster1965}, that is no longer true in the context of parameter estimation or composite hypothesis tests with nuisance parameters. As noted in ~\citek{Whiteson:2006ws}, ``such methods are suboptimal because they assume that the selector with the highest classification accuracy will yield a mass measurement with the smallest statistical uncertainty.''
%In what follows, it will be important to distinguish between the per-event loss used for classification, and the per-experiment loss 
A fundamental distinction is that evaluating the loss for classification is composed of many per-event operations, while evaluating the loss for a mass measurement (e.g. the variance of an estimator for the mass parameter) is a per-experiment operation involving a data set with many events. They went on to demonstrate a computationally intensive stochastic optimization technique based on the per-experiment loss out performed the two stage selection-estimation process. 
%This paper shows that (to the extent a generalized likelihood ratio test in the full feature space is close to  optimal) one can achieve optimal results with a  specific per-event procedure by reformulating the target likelihood ratio in terms of a per-event classification followed by an specific calibration process.

In the context of machine learning or decision theory one starts with a well defined loss function.
In order to define a Bayes risk one must also specify a prior over the parameters.
%In practice, it can be difficult, particularly for large collaborations, to agree on such high-level topics.
In practice, large particle physics collaborations have preferred to avoid specifying the 
high-level loss function and prior, and instead accepted the properties of generalized likelihood ratio tests.
Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself.

The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss. 
This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process. Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.
In Section~\ref{S:Related} we discuss the scheme sketched by \cite{Neal:2007zz} that also suggests using a classifier  as a dimensionality reduction map to aid in the estimation of the likelihood function.

%\bigskip
%
%Generalized likelihood ratio tests enjoy a number of optimal properties in the asymptotic limit and are widely used even away from that limit. These asymptotic properties, the difficulty in specifying a prior over the parameters needed to evaluate a Bayes risk, and the vast literature associated to likelihood ratio tests has led many users of such tests to not be explicit about the loss function they wish to minimize for their experiment. Thus, we will not try to specify a loss function other than to approximate the generalized likelihood ratio itself. The form of the generalized likelihood ratio is explicitly a per-experiment object, thus naively it has the same computational challenges as other approaches that optimize a per-experiment loss.

%Similarly,~\cite{Neal:2007zz} suggested using a classifier as a dimensionality reduction map to aid in the estimation of the likelihood function. We discuss the key differences with that work in Section~\ref{S:Related}.

%This paper shows how one can reformulate generalized likelihood ratio test in terms of a per-event classification stage and a calibration process. Moreover, one can utilize a generative statistical model for the data both to train a discriminative classifier with a supervised learning algorithm and for the calibration process.

@@

We use the following notation:
\begin{itemize}
 \item $x$: a vector of features for an event
 \item $D$: a data set of $D=\{x_1, \dots, x_n\}$, where $x_e$ are assumed to be i.i.d.
 \item $\theta$: parameters of a statistical model
\item $p(x| \theta)$:  probability density  (statistical model) for $x$ given $\theta$
%\item $s(x)$: real-valued score from a machine learning classification algorithm (or any map $s: X\to\mathbb{R}$)
\item $y$: a class label used for training a classifier.
\item $s(x;\theta_0, \theta_1)$: real-valued discriminative classification score, parametrized by $\theta_0$ and $\theta_1$
%\item $p( s | \theta )$ The probability density function for $s$ implied by $p(x|\theta)$ and $s(x)$
\item $p( s_{\theta_0, \theta_1} | \theta )$: The probability density  for $s(x; \theta_0, \theta_1)$ implied by $p(x|\theta)$ 
\end{itemize}
We will assume the $x_e$ are i.i.d., so that $p(D|\theta) = \prod_{e=1}^n p(x_e | \theta)$.

@@

In the setting where one is interested in simple hypothesis testing between a null $\theta=\theta_0$ against an alternate $\theta=\theta_1$, the Neyman-Pearson lemma states that the likelihood ratio 
\begin{equation}
T(D) = \prod_{e=1}^n \frac{ p(x_e|\theta_0)}{ p(x_e|\theta_1)}
\end{equation}
is the most powerful test statistic. In order to evaluate $T(D)$, one must be able to evaluate the probability density 
$p(x| \theta)$ at any value $x$. However, it is increasingly common in science that one has a complex simulation that 
can act as generative model  for $p(x|\theta)$, but one cannot evaluate the density directly. For instance, this is the case 
high energy physics where the simulation of particle detectors can only be done in the `forward mode'. This same setting has been considered by \cite{ClaytonScott}, \cite{JMLR:v14:tong13a}, and \cite{Neal:2007zz}.

@@

While the original goal for frequentist hypothesis testing is to make a decision to accept or reject the null hypothesis based on the entire data set $D$, we are able to reformulate it such that the machine learning problem is a per-event classification problem. This follows from the fact that we assume the $x_e$ to be i.i.d.

@@

Vast literature exists around generative and discriminative classifiers~\citek{AndrewY.Ng}. Typically, generative classifiers learn a model for the joint probability $p(x,y)$, of the inputs $x$ and the classification label $y$, and predict $p(y|x)$ via Bayes rule. In contrast, discriminative classifiers model the posterior $p(y|x)$ directly. For classification tasks, one then thresholds on $p(y|x)$. In both cases this description in terms of a posterior requires a prior distribution for $p(y)$, which is either modeled explicitly or learned from the training data. 
This familiar formulation of classification may lead to some confusion in the setting of the current work.

The first possible source of confusion we wish to avoid is that here $p(x|\theta)$  is a generative \textit{statistical model} for the features $x$, not a generative classifier. We think of the  $p(x|\theta)$ along the lines of a traditional scientific theory or simulator, able to make predictions about $x$ and being motivated by domain-specific considerations. For example, in the context of HEP $p(x|\theta)$ is based on quantum field theory, a detailed simulation of the particle detector, and data processing algorithms that transform raw sensor data into the feature vector $x$~\citek{Sjostrand:2006za,Agostinelli:2002hh}.  
Moreover, we are not attempting to learn the generative model $p(x|\theta)$, we are taking it as given and trying to learn the corresponding (generalized) likelihood ratio test.
%The point here is to make the distinction of $p(x|\theta)$ in this context from a generic generative classifier like normal discriminate analysis or naive Bayes.

The second possible source of confusion is that 
we are not directly interested in calibrating the classification score in terms of a per-event posterior probability $p(y|x)$. 
Instead, we are interested in the approximation of the per-experiment likelihood ratio, which can be used to calculate p-values defined by $P(T(D) > k |\theta)$.
%
%%the likelihood ratio $T(D)$ is aimed at tests based on the entire data set; 
%we are not interested in thresholded classification on individual events $x_e$.  
%Knowing that classifier scores are often poorly calibrated is an immediate concern.

@@

Lastly, in the setting of frequentist hypothesis tests, we do not have a prior $\pi(\theta)$. 
While we can use the generative models to produce training data $(x,y=0)$ generated 
from $p(x|\theta_0)$ and $(x,y=1)$ generated from $p(x|\theta_1)$, the relative mix $p(y)$ 
is arbitrary.  When $p(y=0)=p(y=1)=\sfrac{1}{2}$, then 
\begin{equation}
p(y=1 | x) = \frac{p(x|y=1)}{p(x|y=0)+p(x|y=1)} = \frac{p(x|\theta_1)}{p(x|\theta_0)+p(x|\theta_1))} \;,
\end{equation}
which is monotonic with the desired likelihood ratio $p(x|\theta_1)/p(x|\theta_0)$.
Since the prior $p(y)$ is not needed for the target likelihood ratio test and because the classifier score $p(y|x)$ may not be well calibrated, we choose to denote the classifier score $s(x)$ and simply think of it as a deterministic dimensionality reduction map $s: X \to \mathbb{R}$.  Similar points have been made by~\cite{ClaytonScott} and \cite{Neal:2007zz}.

@@

It is not obvious that if we are working with the distributions $p(s|\theta)$ (for some particular $s(x; \theta_0, \theta_1)$ comparison) that we can find the same estimators. 
Fortunately, there is a construction based on $p(s|\theta)$ that works. The maximum likelihood estimate of Eq.~\ref{eq:mle} is the same as the value that maximizes the likelihood ratio with respect to $p(D|\theta_1)$ for some fixed value of $\theta_1$. This allows us to use Theorem~1 to reformulate the maximum likelihood estimate
\begin{equation}\label{eq:mle_withs}
\hat{\theta} = %\argmax_\theta \frac{ p(D | \theta)}{ p(D | \theta_1)} = 
\argmax_\theta  \sum \ln \frac{p(x_e | \theta)}{p(x_e|\theta_1)} = \argmax_\theta  \sum \ln \frac{p(s(x_e; \theta, \theta_1) | \theta)}{p(s(x_e; \theta, \theta_1) |\theta_1)} \; .
\end{equation}
It is important that we include the denominator $p(s(x_e; \theta, \theta_1) |\theta_1)$ because this cancels Jacobian factors that vary with $\theta$.

@@

We are left with the practical question of how to train a family of discriminative classifiers parametrized by $\theta_0$ and $\theta_1$, the 
parameters associated to the null and alternate hypotheses, respectively. While this could be done independently
for all $\theta_0$ and $\theta_1$, it is desirable and convenient to have a smooth evolution of the classification score as a function of the parameters. Thus, we anticipate a single learning stage based on training data with input $(x, \theta_0, \theta_1)_i$ and target $y_i$. Somewhat unusually, the unknown values of the parameters are taken
as input to the classifier, and are latent variables whose values will be specified via the enveloping (generalized) likelihood ratio test. We denote the learned family of classifiers $\hat{s}(x; \theta_0, \theta_0)$, and anticipate the training based roughly on the following algorithmic flow.
\begin{algorithm}[ht]
\caption{Training of the parametrized classifier.}\label{alg:training}
\begin{algorithmic}
\STATE initialize trainingData to \{\}
\FOR{ $\theta_0$ in $\Theta$ }
	\FOR{ $\theta_1$ in $\Theta$ }
		\STATE generate $x_i^0 \sim p(x|\theta_0)$
		\STATE append $\{ (x_i^0, \theta_0, \theta_1, y=0) \}$ to trainingData
		\STATE generate $x_i^1 \sim p(x|\theta_1)$
		\STATE append $\{ (x_i^1, \theta_0, \theta_1, y=1) \}$ to trainingData
	\ENDFOR
\ENDFOR
\STATE use trainingData to learn $\hat{s}(x; \theta_0, \theta_1)$
\end{algorithmic}
\end{algorithm}%\vspace{-1.5em}

@@

In most settings that make use of likelihood ratio tests, the likelihood is based directly on some approximation of density for the observed data via $\hat{f}(x|\theta)$.  In general, approximating the density $\hat{f}(x|\theta)$ is difficult for high-dimensional data, which motivates the use of the dimensionality reduction map $\hat{s}(x)$ and likelihood ratio tests based on the univariate density $\hat{g}(\hat{s}|\theta)$.  In the case of a fixed classifier $\hat s(x)$ it is possible to pre-compute $\hat s_e=\hat s(x_e)$ and never refer back to the original features $x_e$. In the parametrized setting it is not possible to pre-compute $\hat s(x_e; \theta_0, \theta_1)$ for all values of $\theta_0$ and $\theta_1$; however, we can  embed the classifier into the likelihood function to carry out the composition $\hat{g}\circ \hat{s}$. A concrete realization of this has been performed for probability models implemented with the \texttt{RooFit} probabilistic programing language and  classifiers implemented with \texttt{scikit-learn} and \texttt{TMVA}~\citek{Verkerke:2003ir,scikit-learn,Hocker:2007ht}.

@@

In HEP we are often searching for some 
class of events, generically referred to as \textit{signal}, in the presence of a separate class 
of \textit{background} events.  For each event we measure some quantities $x$ that have corresponding distributions 
$p_b(x|\nu)$ for background and $p_s(x|\nu)$ for signal, where $\nu$ are nuisance parameters describing 
uncertainties in the underlying physics prediction or response of the measurement device. The 
total model is a mixture of the signal and background, and $\mu$ is the mixture coefficient associate 
to the signal component. 
\begin{equation}\label{eq:hepGen}
p( D \,|\, \mu, \nu) = \prod_{e=1}^n \, \left[\, \mu p_s( x_e \, |\,  \nu)  + (1-\mu)\, p_b( x_e \,|\, \nu) \,\right] \; ,
\end{equation}
New particle searches at the LHC are typically framed as hypothesis test where the null corresponds to $\mu=0$, and the
generalized likelihood ratio (profiling over $\nu$ as in Eq.~\ref{eq:cmle}) is used as a test statistic~\citek{Cowan:2010js}. This approach was used for the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.

%Generalized likelihood ratio tests are used widely in HEP ~\citek{Cowan:2010js}, most notably in the discovery of the Higgs boson~\citek{Aad:2012tfa,Chatrchyan:2012ufa}.

@@

This formalism represents is a significant step forward in the usage of machine learning in HEP, where classifiers have always been used between two static classes of events and not parametrized explicitly in terms of the physical quantities we wish to measure. The work of  ~\citek{Whiteson:2006ws} is the closest example as the stochastic optimization was directly trying to minimize the measurement uncertainty of a particle's mass; however, the resulting classifier was fixed. This approach also offers the advantage that it explicitly reformulates the per-experiment optimization to the per-event optimization, which is less computationally intensive.

Another approach that is similar in spirit is the so-called matrix element method, in which one  directly computes an approximate likelihood ratio by performing a computationally intensive integral associated to the detector response~\citek{Volobouev:2011vb}. In the approach considered in this paper, the detector response is naturally handled by the Monte Carlo sampling used in the simulation of the detector; however, that integral is intractable for the matrix element method. Even with drastic simplifications of the detector response, the matrix element method can take several minutes of CPU time per event to calculate the likelihood ratio $q(x)$. The work here can be seen as aiming at the same conceptual target, but utilizing machine learning to overcome the complexity of the detector simulation. It also offers enormous speed increase for evaluating the likelihood at the cost of an initial training stage. In practice, the matrix element method has only been used for searches and measurement of a single physical parameter (sometimes with a single nuisance parameter as in~\citek{Aaltonen:2010yz}).

Contemporary examples where the technique presented here could have major impact on HEP include the measurement of coefficients to quantum mechanical operators describing the decay of the Higgs boson~\citek{Chen:2014pia} and, if we are so lucky, measurement of the mass of supersymmetric particles in cascade decays~\citek{Allanach:2000kt}.  Both of these examples involve data sets with many events, each with a feature vector $x$ that has on the order of 10 components, and a parameter vector $\theta$ with 5-10 parameters of interest and possibly many more nuisance parameters. 
%The state of the art approaches for supersymmetric particle masses typically involve high-level features engineered based on physics intuition
%Many approaches have been tried for measuring supersymmetric particle masses, often based on event level features engineered based on physics intuition and varying 
The state of the art for the operator coefficients of the Higgs decay uses the so-called matrix element likelihood analysis (MELA) in which the equivalent of $s(x; \theta_0, \theta_1)$ is approximated by neglecting detector effects~\citek{Bolognesi:2012mm}.

\subsection{Related work}\label{S:Related}

%Refs to include:
%\citek{ClaytonScott,AndrewY.Ng,SamT.Roweis,EricP.Xing,SujaySanghavi,BiancaZadrozny,Zadrozny2001,}
